\textbf{Effective application of machine learning to bacterial 16S rRNA gene sequencing data}

### 2019-10-03 submission/manuscript.tex
jennas edits

2. Department of Computer Science and Engineering, University or
Michigan, Ann Arbor, MI 49109

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training methods call the validity of these models into question.
Furthermore, there appears to be a preference by many researchers to
favor increased model complexity over interpretability. To overcome
these challenges, we trained seven models that used fecal 16S rRNA
sequence data to predict the presence of colonic screen relevant
neoplasias (SRNs; n=490 patients, 261 controls and 229 cases). We
developed a generalizable pipeline to train, validate and interpret the
models. To show the effect of model selection, we assessed the
predictive performance, interpretability, and computational efficiency
of L2-regularized logistic regression, L1 and L2-regularized support
vector machines (SVM) with linear and radial basis function kernels, a
decision tree, random forest, and extreme gradient boosting (XGBoost).
The random forest model performed best at detecting SRNs with an AUROC
of 0.695 but it was slow to train (83.2 h) and hard to interpret.
Despite its simplicity, L2-regularized logistic regression followed
random forest in predictive performance with an AUROC of 0.680, trained
much faster (12 min), and was more interpretable. Our analysis showed
that ML models should be chosen based on the goal of the study as the
choice will inform expectations of performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable between individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models are useful for understanding the variation in the structure of
existing data and to apply that knowledge to make predictions about new
data. Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual with high confidence
relies on a model that is built with rigorous methods. However, there
are common methodological problems across many of these studies that
need to be addressed as the field progresses. These include a lack of
transparency in which methods are used and how these methods are
implemented; developing and evaluating models without a separate
held-out test data; large variation between the predictive performance
on different folds of cross-validation; and large variation between
cross-validation and testing performances. Nevertheless, the microbiome
field is making progress to avoid some of these pitfalls including
validating their models on independent datasets (7, 18, 19) and
introducing ways to better use ML tools (20--23). More work is needed to
further improve reproducibility and minimize over-optimism for model
performance.

Among microbiome researchers, the lack of transparency in justifying a
modelling approach has been due to an implicit assumption that more
complex models are better because they are more complex. This has
resulted in a trend towards using models such as random forest and
neural networks (2, 11, 24--26) over simpler models such as logistic
regression or other linear models (18, 22, 27). Although the more
complex models may be better at incorporating non-linear relationships
or yield better predictions, they are considered to be black box models
because they are not inherently interpretable. These models require post
hoc explanations to quantify the importance of each feature in making a
prediction and they do not show the structure of how the features are
used. Depending on the application of the model, researchers may choose
to use different modeling approaches. For example, researchers trying to
identify the populations causing a disease would likely want a more
interpretable model whereas clinicians may emphasize predictive
performance. Although one may feel that they are sacrificing
interpretability for performance, that tradeoff may be minimal (28, 29).
Regardless, it is important for researchers to articulate why they have
selected a specific modelling approach or even compare multiple
approaches in the same study.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing 7 modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also built four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared their
predictive performance, interpretability, and computational efficiency.
To demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classifiy individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (3). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (1, 3, 4, 30). With this dataset we developed a framework
that implements a ML pipeline that can be used for any modeling
approach, evaluates predictive performance, and demonstrates how to
interpret these models. This framework can be easily applied to other
host-associated and environmental microbiome datasets.

@@

\textbf{Model selection and pipeline construction} We established a
common ML pipeline to train and validate each of the seven models that
is based on standard methods within the ML community (REFS by
Jenna?){[}Figure 1{]}.

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset while the test
set was composed of the remaining 20\% of the data {[}Figure 1{]}. To
maintain the distribution of controls and cases that was found with the
full dataset, we performed stratified splits. For example, our full
dataset included 490 individuals. Of these, 261 had normal colons (53\%)
and 229 had a screen relevant neoplasia (SRN; 46.7\%). A training set
included 393 individuals, of which 209 had an SRN (53\%), while the test
set was composed of 97 individuals of which 52 had an SRN (54\%). The
training data was used to build the models and the test set was used for
evaluating predictive performance.

Second, we trained seven different models using the training data
{[}Table 1{]}. We selected models with different classification
algorithms and regularization methods. Regularization is a technique
that discourages overfitting by penalizing the model for learning the
training data too well. For regularized logistic regression and SVM with
linear kernel, we used L2-regularization to keep all potentially
important features. For comparison, we also trained an L1-regularized
SVM model with linear kernel. L1-regularization on microbiome data led
to a sparser solution (i.e., forced many coefficients to zero). To
explore the potential for non-linear relationships among features to
improve classification, we trained tree-based models including decision
tree, random forest, and XGBoost and we trained an SVM model with a
non-linear kernel.

Third, fitting of these models required selecting appropriate
hyperparameters. Hyperparameters are the rules that are learned from the
training set in a classification algorithm. For example, in the linear
models the regularization term (C) is a hyperparameter that indicates
the penalty for overfitting. Similar to regularization term C, all
hyperparameters are tuned to find the best model. We selected
hyperparameters by performing 100 five-fold cross-validation (CV)
repeats on the training set {[}Figure 1{]}. The five-fold CV was also
stratified to maintain the overall case and control distribution. We
chose the best hyperparameter values for each model based on its CV
predictive performance using the area under the receiver operating
characteristic curve (AUROC) metric {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0.50, where the model's predictions are no different
from random chance. To select the best performing hyperparameter, we
performed a full grid search for hyperparameter settings when training
our models. Default hyperparameter settings in previously developed ML
packages in R, Python, and Matlab programming languages are inadequate
for effective application of classification algorithms and need to be
optimized for each new ML task. In the example of L1-regularized SVM
with linear kernel {[}Figure S1{]}, the model showed large variability
between different regularization coefficients (C) and was susceptible to
performing poorly if the wrong regularization coefficient was assigned
to the model by default.

Finally, we trained the full training dataset with the selected
hyperparameter values and applied the model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to obtain a reliable and robust interpretation of
model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and SVM
with linear kernel) performed nearly as well as random forest.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
would indicate that the models were overfit. The difference in median
AUROCs was 0.021 in L1-regularized SVM with linear kernel, followed by
SVM with radial basis function kernel and decision tree with a
difference of 0.007 and 0.006, respectively {[}Figure 2{]}. These
differences are relatively small and would not indicate a problem with
overfitting.

To evaluate the risk for over-optimism of each model, we calculated the
range of AUROC values for each model using 100 data-splits. The range
among the testing AUROC values within each model varied by 0.230 on
average across the seven models. If we had only done a single split,
then there is a risk that we could gotten lucky or unlucky with the
performance of the model. For instance, the lowest AUROC value of the
random forest model was 0.593 whereas the highest was 0.810. These
results showed that depending on the data-split, the testing AUROC
values showed great variability {[}Figure 2{]}. Therefore, it is
important to employ the 100 hierarchical data splits that were included
in our pipeline to minimize the risk of over-optimism.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
range among the cross-validation AUROC values within both models at
lower sample sizes were much larger than when the full collection of
samples was used to train and validate the models. These results showed
that because the microbiome data had many features (6920 OTUs), it was
important to train the models using appropriate sample sizes to avoid
problems with generalizability and over-optimism. Furthermore, it was
encouraging that even for a small number of samples, the interquartile
range included the median AUROC values for the larger subsetted
datasets.

\textbf{Interpretation of each ML model.} Interpretability is the degree
to which humans can understand the reasons behind a model prediction
(31). Because we often use ML models not just to predict a health
outcome but also to identify the biomarkers for a disease, model
interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

@@

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a posthoc explanation of the model where we randomly permuted
non-correlated features individually and groups of perfectly correlated
features across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e testing
AUROC values) decreased when each OTU or group of OTUs was randomly
permuted. We ranked the OTUs based on how much the median testing AUROC
decreased when it was permuted; the OTU with the largest decrease ranked
highest {[}Figure 4{]}. Among the twenty OTUs with the largest impact,
there was only one OTU (OTU 822) that was shared among all of the
models; however, we found three OTUs (OTU 58, 110, 367) that were
important in each of the tree-based models. Similarly, the random forest
and XGBoost models, shared four of the most important OTUs (OTU 2, 12,
361, 477). Permutation analysis results also revealed that with the
exception of the decision tree model, removal of any individual OTU had
a minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
Effectively, the complexity of the communities was more fully
represented in the better performing models (22, 32). At least in this
case, it was not possible to distinguish between health and disease
using a single OTU. Although permutation analysis allowed us to gauge
the importance of an OTU, the analysis was post-hoc (i.e.~done using the
test data) and these results did not allow us to directly interrogate
the models to know whether an OTU was associated with classifying a
subject as being healthy or having an SRN.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. As expected, the training times
increased with the complexity of the model and the number of tuned
hyperparameter settings. Also, the linear models trained faster than
non-linear models {[}Figures S1-S2; Figure 5{]}. When we subsetted the
size of the training dataset, we observed a linear relationship between
the size of the dataset and the training times for L2-regularized
logistic regression and random forest models {[}Figure S5{]}.

@@

There is a growing awareness that many human diseases and environmental
processes are not driven by a single organism but are the product of
multiple bacterial populations. Traditional statistical approaches are
useful for identifying those cases where a single organism is associated
with a process. In contrast, ML methods offer the ability to incorporate
the structure of the microbial communities as a whole to classify them
into different categories such as coming from a patient who is healthy
or has SRNs. If it is possible to classify communities reliably, then ML
methods also offer the ability to identify those microbial populations
within the communities that are responsible for the classification.
However, the application of ML in microbiome studies is still in its
infancy and the field needs to develop a better understanding of
different ML methods, their strengths and weaknesses, and how to
implement them.

To address these needs, we developed a framework to train rigorous,
transparent, and reproducible models. We benchmarked seven ML models and
showed that we can create models that are inherently interpretable and
easily trained without losing predictive performance. In terms of
predictive performance, the random forest model had the best testing
AUROC values compared to the other six models. However, the second-best
model was L2-regularized logistic regression with a median AUROC
difference of only 0.015 compared to random forest. While random forest
took 83.2 hours to train, L2-regularized logistic regression trained in
12 minutes. In terms of interpretability, random forest is a complex ML
model and could only be explained using post-hoc methods such as
permutation importance. On the other hand, L2-regularized logistic
regression was easier to interpret by ranking the OTUs based on their
feature weights and reviewing the signs of these weights. Comparing many
different models showed us that the most complex model was not
necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that
can be generalized to any modeling method that uses 16S rRNA sequence
counts to predict a binary health outcome. We performed a random
data-split to create a training set (80\% of the data) and a held-out
test set (20\% of the data), which we used to evaluate predictive
performance. We repeated this data-split 100 times to measure the
possible variation in predictive performance. During the training, we
tuned the model hyperparameters with a repeated five-fold
cross-validation. Despite the high number of features microbiome
datasets typically have, the models we built with this pipeline were
generalizable as shown by the similar AUROC values from the
cross-validation and testing.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (33). For example if one
OTU is highly correlated with another OTU, only one of these OTUs would
have a large feature weight thus hiding the importance of its correlated
OTU. To capture all important features, once we identify highly ranked
OTUs, we should review their relationships with other OTUs. These
relationships will help us generate new hypotheses about the ecology of
the disease and test them with follow-up experiments. When we used
permutation importance, we took collinearity into consideration by
grouping correlated OTUs to determine their impact as a group. We
grouped OTUs that had a perfect correlation with each other however, we
can reduce the correlation threshold to further investigate the
relationships among features. It is important to know the correlation
structures of the data to avoid misinterpreting the models. This is
likely to be a particular problem with shotgun metagenomic datasets
where collinearity will be more pronounced due to many genes being
correlated with one another because they come from the same chromosome.
To identify the true underlying microbial factors of a disease, it is
crucial to do correlation analyses and further experimentation for
biological validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as number of categories to classify increase, more samples
are required for each category to train a model. We also did not use
regression-based analyses to predict a non-categorical outcome. We have
previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (34). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold great promise (11, 35, 36) but microbiome datasets
often suffer from having many features and small sample sizes, which
makes the deep learning models prone to overfitting. These methods are
even more complex than random forest and XGBoost and are considered
uninterpretable. There is great potential for applying ML approaches to
microbiome data and we can choose which ML approach is best for the
study.

Our framework gives structure to investigators wanting to train,
evaluate, and interpret their own ML models to identify OTUs that might
be biologically relevant. However, deploying microbiome-based models to
make clinical diagnoses or predictions is a significantly harder and
distinct undertaking. For example, we currently lack standardized
methods to collect patient samples, generate sequence data, and report
clinical data. We are also challenged by the practical constraints of
OTU-based approaches. The de novo algorithms commonly in use are slow,
require considerable memory, and result in different OTU assignments as
new data are added. Finally, we also need independent validation cohorts
to test the performance of a diagnostic model. To realize the potential
for using ML approaches with microbiome data, it is necessary that we
direct our efforts to overcome these challenges.

>>>>

2. Department of Electrical Engineering and Computer Science, University
or Michigan, Ann Arbor, MI 48109

@@

Machine learning (ML) modeling of the human microbiome has the potential
to identify microbial biomarkers and aid in diagnosis of many diseases
such as inflammatory bowel disease, diabetes, and colorectal cancer.
Progress has been made towards developing ML models that predict health
outcomes using bacterial abundances, but inconsistent adoption of
training and evaluation methods call the validity of these models into
question. Furthermore, there appears to be a preference by many
researchers to favor increased model complexity over interpretability.
To overcome these challenges, we trained seven models that used fecal
16S rRNA sequence data to predict the presence of colonic screen
relevant neoplasias (SRNs; n=490 patients, 261 controls and 229 cases).
We developed a generalizable pipeline to train, validate and interpret
the models. To show the effect of model selection, we assessed the
predictive performance, interpretability, and training time of
L2-regularized logistic regression, L1 and L2-regularized support vector
machines (SVM) with linear and radial basis function kernels, decision
trees, random forest, and extreme gradient boosted trees (XGBoost). The
random forest model performed best at detecting SRNs with an AUROC of
0.695 {[}IQR 0.651-0.739{]} but was slow to train (83.2 h) and difficult
to interpret. Despite its simplicity, L2-regularized logistic regression
followed random forest in predictive performance with an AUROC of 0.680
{[}IQR 0.625-0.735{]}, trained faster (12 min), and was more
interpretable. Our analysis showed that ML models should be chosen based
on the goal of the study, as the choice will inform expectations of
performance and interpretability.

@@

As the number of people represented in human microbiome datasets grow,
there is an increasing desire to use microbiome data to diagnose
diseases. However, the structure of the human microbiome is remarkably
variable AMONG individuals to the point where it is often difficult to
identify the bacterial populations that are associated with diseases
using traditional statistical models. This variation is likely due to
the ability of many bacterial populations to fill the same niche such
that different populations cause the same disease in different
individuals. Furthermore, a growing number of studies have shown that it
is rare for a single bacterial species to be associated with a disease.
Instead, subsets of the microbiome account for differences in health.
Traditional statistical approaches do not adequately account for the
variation in the human microbiome and typically consider the protective
or risk effects of each bacterial population individually. Recently,
machine learning models (ML) have grown in popularity among microbiome
researchers because of the large amount of data that can now be
generated and because the models are effective at accounting for the
interpersonal microbiome variation and the ecology of the disease.

ML models can be used to increase our understanding of the variation in
the structure of existing data and in making predictions about new data.
Researchers have used ML models to diagnose and understand the
ecological basis of diseases such as liver cirrhosis, colorectal cancer,
inflammatory bowel diseases, obesity, and type 2 diabetes (1--16,
16--18). The task of diagnosing an individual relies on a rigorously
validated model. However, there are common methodological and reporting
problems that arise when applying ML to such data, that need to be
addressed for the field to progress. These problems include a lack of
transparency in which methods are used and how these methods are
implemented; evaluating models without a separate held-out test data;
large variation between the predictive performance on different folds of
cross-validation; and large variation between cross-validation and
testing performances. Nevertheless, the microbiome field is making
progress to avoid some of these pitfalls including validating their
models on independent datasets (7, 18, 19) and introducing ways to
better use ML tools (20--23). More work is needed to further improve
reproducibility and minimize overestimating for model performance.

Among microbiome researchers, the lack of justification when selecting a
modelling approach has been due to an implicit assumption that more
complex models are better. This has resulted in a trend towards using
models such as random forest and deep neural networks (2, 11, 24--26)
over simpler models such as logistic regression or other linear models
(18, 22, 27). Although complex models may in some cases, capture
important non-linear relationships and therefore yield better
predictions, they may result in black boxes that lack interpretibility.
Such models require post hoc explanations to quantify the importance of
each feature in making predictions. Depending on the goal of the
modeling, researchers may choose to use different approaches. For
example, researchers trying to identify the populations of microbiota
associated wwith disease may desire a more interpretable model, whereas
clinicians may emphasize predictive performance. Nonetheless, it is
important to understand the tradeoff and accepting that this tradeoff
may be minimal (28, 29). It is important for researchers to justify
their choice of modelling approach.

To showcase a rigorous ML pipeline and to shed light on how ML model
selection can affect modeling results, we performed an empirical
analysis comparing seven modeling approaches with the same dataset and
pipeline. We built three linear models with different forms of
regularization: L2-regularized logistic regression and L1 and
L2-regularized support vector machines (SVM) with a linear kernel. We
also trained four non-linear models: SVM with radial basis function
kernel, a decision tree, random forest and XGBoost. We compared their
predictive performance, interpretability, and training time. To
demonstrate the performance of these modeling approaches and our
pipeline, we used data from a previously published study that sought to
classifiy individuals as having normal colons or colonic lesions based
on the 16S rRNA gene sequences collected from fecal samples (3). This
dataset was selected because it is a relatively large collection of
individuals (N=490) connected to a clinically significant disease where
there is ample evidence that the disease is driven by variation in the
microbiome (1, 3, 4, 30). With this dataset we developed a ML pipeline
that can be used in many different scenarios for training and evaluating
models. This framework can be easily applied to other host-associated
and environmental microbiome datasets.

@@

\textbf{Model selection and pipeline construction} We established a
reusable ML pipeline for model selection and evaluation, focusing on
seven different commonly used supervised learning algorithms {[}Figure
1{]}.

First, we randomly split the data into training and test sets so that
the training set consisted of 80\% of the full dataset, while the test
set was composed of the remaining 20\% {[}Figure 1{]}. To maintain the
distribution of controls and cases found in the full dataset, we
performed stratified splits. For example, our full dataset included 490
individuals. Of these, 261 had normal colons (53\%) and 229 had a screen
relevant neoplasia (SRN; 46.7\%). A training set included 393
individuals, of which 209 had an SRN (53\%), while the test set was
composed of 97 individuals of which 52 had an SRN (54\%). The training
data was used to build and select the models and the test set was used
for evaluating the model.

We trained seven different models using the training data {[}Table 1{]}.
We focused on different classification algorithms and regularization
methods. Regularization helps to prevent overfitting by penalizing a
model that fits the training data too well. For regularized logistic
regression and SVM with a linear kernel, we used L2-regularization to
keep all potentially important features. For comparison, we also trained
an L1-regularized SVM with a linear kernel. L1-regularization on
microbiome data led to a sparser solution (i.e., forced many
coefficients to zero). To explore the potential for non-linear
relationships among features to improve classification, we trained
tree-based models including a decision tree, a random forest, and
XGBoost and an SVM with a non-linear kernel.

Model selection required tuning hyperparameters. Hyperparameters are
parameters which need to be specified or tuned by the user, in order to
train a model for a specific modeling problem. For example, when using
regularization, (C) is a hyperparameter that indicates the penalty for
overfitting. Hyperparameters are tuned using the training data to find
the best model. We selected hyperparameters by performing repeated
five-fold cross-validation (CV) on the training set {[}Figure 1{]}. The
five-fold CV was also stratified to maintain the overall case and
control distribution. We chose the hyperparameter values that led to the
best average CV predictive performance using the area under the receiver
operating characteristic curve (AUROC) {[}Figure S1 and S2{]}. The AUROC
ranges from 1.0, where the model perfectly distinguishes between cases
and controls, to 0, where the model's predictions are perfectly
incorrect. An AUROC value of 0.5 indicates that model's predictions are
no differect than random. To select hyperparameters, we performed a full
grid search for hyperparameter settings when training our models.
Default hyperparameter settings in previously developed ML packages in
R, Python, and MATLAB programming languages may be inadequate for
effective application of classification algorithms and need to be
optimized for each new ML task. In the example of L1-regularized SVM
with linear kernel {[}Figure S1{]}, the model showed large variability
between different regularization strengths (C).

Once hyperparameters were selected, we trained the model using the full
training dataset and applied the final model to the held-out data to
evaluate the testing predictive performance of each model. The
data-split, hyperparameter selection, training and testing steps were
repeated 100 times to obtain a reliable and robust interpretation of
model performance {[}Figure 1{]}.

\textbf{Predictive performance and generalizability of the seven
models.} We evaluated the predictive performance of the seven models to
classify individuals as having normal colons or SRNs {[}Figure 2{]}. The
random forest model had significantly higher test AUROC values than the
other models for detecting SRNs (Wilcoxon rank sum test, p \textless{}
0.01). The median AUROC of the random forest model was 0.695 (IQR
0.044). L2-regularized logistic regression, XGBoost, L2-regularized SVM
with linear and radial basis function kernel AUROC values were not
significantly different from one another and had median AUROC values of
0.680 (IQR 0.055), 0.679 (IQR 0.052), 0.678 (IQR 0.056) and 0.668 (IQR
0.056), respectively. L1-regularized SVM with linear kernel and decision
tree had significantly lower AUROC values than the other ML models with
median AUROC of 0.650 (IQR 0.066) and 0.601 (IQR 0.059), respectively
{[}Figure 2{]}. Interestingly, these results demonstrate that the most
complex model (XGBoost) did not have the best performance and that the
most interpretable models (L2-regularized logistic regression and SVM
with linear kernel) performed nearly as well non-linear models.

To evaluate the generalizability of each model, we compared the median
cross-validation AUROC to the median testing AUROC. If the difference
between the cross-validation and testing AUROCs was large, then that
could indicate that the models were overfit to the training data. The
difference in median AUROCs was 0.021 in L1-regularized SVM with linear
kernel, followed by SVM with radial basis function kernel and decision
tree with a difference of 0.007 and 0.006, respectively {[}Figure 2{]}.
These differences are relatively small and gives us confidence in our
estimate of the generalization performance of the models.

To evaluate the variation in the estimated performance, we calculated
the range of AUROC values for each model using 100 data-splits. The
range among the testing AUROC values within each model varied by 0.230
on average across the seven models. If we had only done a single split,
then there is a risk that we could have gotten lucky or unlucky in
estimating model performance. For instance, the lowest AUROC value of
the random forest model was 0.593 whereas the highest was 0.810. These
results showed that depending on the data-split, the testing performance
can vary {[}Figure 2{]}. Therefore, it is important to employ multiple
data splits when estimating generalization performance.

To show the effect of sample size on model generalizability, we compared
cross-validation AUROC values of L2-regularized logistic regression and
random forest models when we subsetted our original study design with
490 subjects to 15, 30, 60, 120, and 245 subjects {[}Figure S3{]}. The
variation in cross-validation performance within both models at lower
sample sizes was larger than when the full collection of samples was
used to train and validate the models. Because of the high
dimensionality of the microbiome data (6920 OTUs), large sample sizes
can lead to better models.

\textbf{Interpretation of each ML model.} Interpretability is related to
the degree to which humans can understand the reasons behind a model
prediction (31--33). Because we often use ML models not just to predict
a health outcome but also to identify the biomarkers for a disease,
model interpretation becomes crucial for microbiome studies. ML models
decrease in interpretability as they increase in complexity. In this
study we used two methods to help interpret our models.

@@

Second, to analyze non-linear models we interpreted the feature
importance using permutation importance. Whereas the absolute feature
weights were determined from the trained models, here we measured
importance using the held-out test data. Permutation importance analysis
is a posthoc explanation of the model, in which we randomly permuted
groups of perfectly correlated features together and other features
individually across the two groups in the held-out test data. We then
calculated how much the predictive performance of the model (i.e,
testing AUROC values) decreased when each OTU or group of OTUs was
randomly permuted. We ranked the OTUs based on how much the median
testing AUROC decreased when it was permuted; the OTU with the largest
decrease ranked highest {[}Figure 4{]}. Among the twenty OTUs with the
largest impact, there was only one OTU (OTU 822) that was shared among
all of the models; however, we found three OTUs (OTU 58, 110, 367) that
were important in each of the tree-based models. Similarly, the random
forest and XGBoost models, shared four of the most important OTUs (OTU
2, 12, 361, 477). Permutation analysis results also revealed that with
the exception of the decision tree model, removal of any individual OTU
had minimal impact on model performance. For example, if OTU 367 was
permuted across the samples in the decision tree model, the median AUROC
dropped from 0.601 to 0.525. In contrast, if the same OTU was permuted
in the random forest model, the AUROC only dropped from 0.695 to 0.680.
In this case, similar to previous studies (22, 34), it was not possible
to distinguish between health and disease using a single OTU. Although
permutation analysis allowed us to gauge the importance of an OTU, the
analysis was post-hoc (i.e.~done using the test data) and these results
did not allow us to directly interrogate the models to know whether an
OTU was associated with classifying a subject as being healthy or having
an SRN.

@@

\textbf{The computational efficiency of each ML model.} We compared the
training times of the seven ML models. As expected, the training times
increased with the complexity of the model and the number of potential
hyperparameter combinations. Also, the linear models trained faster than
non-linear models {[}Figures S1-S2; Figure 5{]}.

@@

There is a growing awareness that many human diseases and environmental
processes are not driven by a single organism but are the product of
multiple bacterial populations. Traditional statistical approaches are
useful for identifying those cases where a single organism is associated
with a process. In contrast, ML methods offer the ability to incorporate
the structure of the microbial communities as a whole and identify
associations between community structure and disease state. If it is
possible to classify communities reliably, then ML methods also offer
the ability to identify those microbial populations within the
communities that are responsible for the classification. However, the
application of ML in microbiome studies is still in its infancy and the
field needs to develop a better understanding of different ML methods,
their strengths and weaknesses, and how to implement them.

To address these needs, we developed an open-sourced framework to ML
models. using this pipeline, we benchmarked seven ML models and showed
that the tradeoff between model complexity and performance may be less
severe than originally hypothesized. In terms of predictive performance,
the random forest model had the best AUROC compared to the other six
models. However, the second-best model was L2-regularized logistic
regression with a median AUROC difference of only 0.015 compared to
random forest. While our implementation of random forest took 83.2 hours
to train, our L2-regularized logistic regression trained in 12 minutes.
In terms of interpretability, random forest is a non-linear ML model and
is most often explained using post-hoc methods such as permutation
importance. On the other hand, using L2-regularized logistic regression
we ranked the importance of each OTU based on their feature weights.
Comparing many different models showed us that the most complex model
was not necessarily the best model for our ML task.

As we set out to select the best model, we established a pipeline that
can be generalized to any modeling method that predicts a binary health
outcome. We performed a random data-split to create a training set (80\%
of the data) and a held-out test set (20\% of the data), which we used
to evaluate predictive performance. We repeated this data-split 100
times to measure the possible variation in predictive performance.
During the training, we tuned the model hyperparameters with a repeated
five-fold cross-validation. Despite the high number of features
microbiome datasets typically have, the models we built with this
pipeline were generalizable as shown by the good test AUROCs.

We highlighted the importance of model interpretation to gain greater
biological insights into microbiota-associated diseases. In this study
we showcased two different interpretation methods: ranking each OTU by
(i) their absolute weights in the trained models and (ii) their impact
on the predictive performance based on permutation importance.
Human-associated microbial communities have complex correlation
structures which create collinearity in the datasets. This can hinder
our ability to reliably interpret models because the feature weights of
correlated OTUs are influenced by one another (35). To capture all
important features, once we identify highly ranked OTUs, we should
review their relationships with other OTUs. These relationships will
help us generate new hypotheses about the ecology of the disease and
test them with follow-up experiments. When we used permutation
importance, we took collinearity into consideration by grouping
correlated OTUs to determine their impact as a group. We grouped OTUs
that had a perfect correlation with each other however, we can reduce
the correlation threshold to further investigate the relationships among
correlated features. It is important to know the correlation structures
of the data to avoid misinterpreting the models. This is likely to be a
particular problem with shotgun metagenomic datasets where collinearity
will be more pronounced due to many genes being correlated with one
another because they come from the same chromosome. To identify the true
underlying microbial factors of a disease, it is crucial to do
correlation analyses and further experimentation for biological
validation.

In this study, we did not consider all possible modeling approaches.
However, the principles highlighted throughout this study apply to other
ML modeling tasks with microbiome data. For example, we did not evaluate
multicategory classification methods to predict non-binary outcomes. We
could have trained models to differentiate between people with normal
colons and those with adenomas or carcinomas (k=3 categories). We did
not perform this analysis because the clinically relevant diagnosis
grouping was between patients with normal colons and those with SRNs.
Furthermore, as number of classes increases, more samples are required
for each category to train an accurate model. We also did not use
regression-based analyses to predict a non-categorical outcome. We have
previously used such an approach to train random forest models to
predict fecal short-chain fatty acid concentrations based on microbiome
data (36). Our analysis was also limited to shallow learning methods and
did not explore deep learning methods such as neural networks. Deep
learning methods hold promise (11, 37, 38) but microbiome datasets often
suffer from having many features and small sample sizes, which can
result in overfitting.

Our framework provides a reproducible structure to investigators wanting
to train, evaluate, and interpret their own ML models to generate
hypotheses regarding which OTUs might be biologically relevant. However,
deploying microbiome-based models to make clinical diagnoses or
predictions is a significantly harder and distinct undertaking. For
example, we currently lack standardized methods to collect patient
samples, generate sequence data, and report clinical data. We are also
challenged by the practical constraints of OTU-based approaches. The de
novo algorithms commonly in use are slow, require considerable memory,
and result in different OTU assignments as new data are added. Finally,
we also need independent validation cohorts to test the performance of a
diagnostic model. To realize the potential for using ML approaches with
microbiome data, it is necessary that we direct our efforts to overcome
these challenges.

Our study highlights the need to make educated choices at every step of
developing a ML model with microbiome data. We created an aspirational
rubric that researchers can use to identify potential pitfalls when
using ML in microbiome studies and ways to avoid them {[}Table S1{]}. We
have highlighted the trade-offs between model complexity and
interpretability, the need for tuning hyperparameters, the utility of
held-out test sets for evaluating predictive performance, and the
importance of considering correlation structures in datasets for
reliable interpretation. Furthermore, we underscored the importance of
proper experimental design and methods to help us achieve the level of
validity and accountability we want from models built for patient
health.
